
# R Graphics

**Topics**

* R `ggplot2` package
* Setup basic plots
* Add and modify scales and legends
* Manipulate plot labels
* Change and create plot themes


## Setup

### Software & materials

You should have R and RStudio installed --- if not:

* Download and install R: <http://cran.r-project.org>
* Download and install RStudio: <https://www.rstudio.com/products/rstudio/download/#download>

Download materials:

* Download class materials at <https://github.com/IQSS/dss-workshops-redux/raw/master/R/Rgraphics.zip>
* Extract materials from the zipped directory `Rgraphics.zip` (Right-click => Extract All on Windows, double-click on Mac) and move them to your desktop!

Start RStudio and create a new project:

* On Windows click the start button and search for RStudio. On Mac
    RStudio will be in your applications folder.
* In Rstudio go to `File -> New Project`.
* Choose `Existing Directory` and browse to the `Rgraphics` directory.
* Choose `File -> Open File` and select the blank version of the `.Rmd` file.

While R's built-in packages are powerful, in recent years there has
been a big surge in well-designed *contributed packages* for R. 
In particular, a collection of R packages called 
[tidyverse](https://www.tidyverse.org/) have been 
designed specifically for data science. All packages included in 
`tidyverse` share an underlying design philosophy, grammar, and 
data structures. We will use `tidyverse` packages throughout the 
workshop, so let's install them now, together with the `scales` and
`ggrepel` packages that provide additional functionality to `ggplot2`.

```{r}
# install.packages("tidyverse")
#

# install.packages("scales")
#

# install.packages("ggrepel") 
#

```

We can also install the `rmarkdown` package, which will allow us to
combine our text and code into a formatted document at the end of 
the workshop:

```{r}
# install.packages("rmarkdown")
#

```

### Goals

Class Structure and Organization:

* Ask questions at any time. Really!
* Collaboration is encouraged - please spend a minute introducing yourself to your neighbors!

This is an intermediate R course:

* Assumes working knowledge of R
* Relatively fast-paced
* Focus is on `ggplot2` graphics; other packages will not be covered


## Why `ggplot2`?

`ggplot2` is a package within in the `tidyverse` suite of packages. Advantages of `ggplot2` include:

* consistent underlying `grammar of graphics` (Wilkinson, 2005)
* plot specification at a high level of abstraction
* very flexible
* theme system for polishing plot appearance
* mature and complete graphics system
* many users, active mailing list

That said, there are some things you cannot (or should not) do with `ggplot2`:

* 3-dimensional graphics (see the `rgl` package)
* Graph-theory type graphs (nodes/edges layout; see the `igraph` package)
* Interactive graphics (see the `ggvis` package)

### What is the Grammar Of Graphics?

The basic idea: independently specify plot building blocks and combine them to create just 
about any kind of graphical display you want. Building blocks of a graph include the
following (**bold denotes essential elements**):

* **data**
* **aesthetic mapping**
* **geometric object**
* statistical transformations
* scales
* coordinate system
* position adjustments
* faceting
* themes


### `ggplot2` VS base graphics

Compared to base graphics, `ggplot2`

* is more verbose for simple / canned graphics
* is less verbose for complex / custom graphics
* does not have methods (data should always be in a `data.frame`)
* uses a different system for adding plot elements
* has sensible defaults for generating legends


## Geometric objects & aesthetics

### Aesthetic mapping

In ggplot land *aesthetic* means "something you can see". Examples include:

* position (i.e., on the x and y axes)
* color ("outside" color)
* fill ("inside" color)
* shape (of points)
* linetype
* size

Each type of geom accepts only a subset of all aesthetics; refer to the geom help pages to see what mappings each geom accepts.
Aesthetic mappings are set with the `aes()` function.

### Geometric objects (`geom`)

Geometric objects are the actual marks we put on a plot. Examples include:

* points (`geom_point()`, for scatter plots, dot plots, etc.)
* lines (`geom_line()`, for time series, trend lines, etc.)
* boxplot (`geom_boxplot()`, for boxplots!)

A plot **must have at least one geom**; there is no upper limit. You can add a geom to a plot using the `+` operator.

Each `geom_` has a particular set of aesthetic mappings associated with it. Some examples are provided below, 
with required aesthetics in **bold** and optional aesthetics in plain text:

| `geom_`          | Usage             | Aesthetics                                                                      |
|:-----------------|:------------------|:-----------------------------------------------------------------------------------------|
| `geom_point()`   | Scatter plot      |**`x`**,**`y`**,`alpha`,`color`,`fill`,`group`,`shape`,`size`,`stroke`                    |
| `geom_line()`    | Line plot         |**`x`**,**`y`**,`alpha`,`color`,`linetype`,`size`                                         |
| `geom_bar()`     | Bar chart         |**`x`**,**`y`**,`alpha`,`color`,`fill`,`group`,`linetype`,`size`                          |
| `geom_boxplot()` | Boxplot           |**`x`**,**`lower`**,**`upper`**,**`middle`**,**`ymin`**,**`ymax`**,`alpha`,`color`,`fill` |
| `geom_density()` | Density plot      |**`x`**,**`y`**,`alpha`,`color`,`fill`,`group`,`linetype`,`size`,`weight`                 |
| `geom_smooth()`  | Conditional means |**`x`**,**`y`**,`alpha`,`color`,`fill`,`group`,`linetype`,`size`,`weight`                 |
| `geom_label()`   | Text              |**`x`**,**`y`**,**`label`**,`alpha`,`angle`,`color`,`family`,`fontface`,`size`            |

You can get a list of all available geometric objects and their associated aesthetics at <https://ggplot2.tidyverse.org/reference/>

or simply type `geom_<tab>` in any good R IDE (such as Rstudio or ESS) to see a list of functions starting with `geom_`.


#### Points (scatterplot)

Now that we know about geometric objects and aesthetic mapping, we can make a `ggplot()`. `geom_point()` requires mappings for x and y, all others are optional.

**Example data: housing prices**

Let's look at housing prices.

```{r}
housing <- read_csv("dataSets/landdata-states.csv")
#

# create a subset for 1st quarter 2001
#

```

**Step 1:** create a blank canvas by specifying data:

```{r} 
#

```

**Step 2:** specify aesthetic mappings (how you want to map variables to visual aspects):

```{r}
# here we map "Land_Value" and "Structure_Cost" to the x- and y-axes.
#

```

**Step 3:** add new layers of geometric objects that will show up on the plot: 

```{r}
# here we use geom_point() to add a layer with point (dot) elements 
# as the geometric shapes to represent the data.
#

```


#### Lines (prediction line)

A plot constructed with `ggplot()` can have more than one geom. In that case the mappings established in the `ggplot()` call are plot defaults that can be added to or overridden. Our plot could use a regression line:

```{r}
#

#

#

```


#### Smoothers

Not all geometric objects are simple shapes; the smooth geom includes a line and a ribbon.

```{r}
#

```


#### Text (label points)

Each geom accepts a particular set of mappings; for example `geom_text()` accepts a `label` mapping.

```{r}
#

```


```{r}
#

```


### Aesthetic mapping VS assignment

Note that variables are mapped to aesthetics with the `aes()` function, while fixed aesthetics are set outside the `aes()` call. 
This sometimes leads to confusion, as in this example:

```{r}
# incorrect! 2 is not a variable
# this is fine -- all points red

```


### Mapping variables to other aesthetics

Other aesthetics are mapped in the same way as x and y in the previous example.

```{r}
#

```


## Exercise 0

The data for the exercises is available in the `dataSets/EconomistData.csv` file. Read it in with

```{r}
dat <- read_csv("dataSets/EconomistData.csv")
```

Original sources for these data are <http://www.transparency.org/content/download/64476/1031428> <http://hdrstats.undp.org/en/indicators/display_cf_xls_indicator.cfm?indicator_id=103106&lang=en>

These data consist of *Human Development Index* and *Corruption Perception Index* scores for several countries.

1.  Create a scatter plot with `CPI` on the x axis and `HDI` on the y axis.
```{r}
## 
```

2.  Color the points in the previous plot blue.
```{r}
## 
```

3.  Map the color of the the points to `Region`.
```{r}
## 
```

4.  Keeping color mapped to `Region`, make the points bigger by setting size to 2
```{r}
## 
```

5.  Keeping color mapped to `Region`, map the size of the points to `HDI_Rank`
```{r}
## 
```


## Statistical transformations

### Why transform data?

Some plot types (such as scatterplots) do not require transformations; each point is plotted at x and y coordinates equal to the original value. Other plots, such as boxplots, histograms, prediction lines etc. require statistical transformations:

* for a boxplot the y values must be transformed to the median and 1.5(IQR)
* for a smoother the y values must be transformed into predicted values

Each geom has a default statistic, but these can be changed. For example, the default statistic for `geom_histogram()` is `stat_bin()`:

```{r}
#

```

Here is a list of geoms and their default statistics <https://ggplot2.tidyverse.org/reference/>


### Setting arguments

Arguments to `stat_` functions can be passed through `geom_` functions. This can be slightly annoying because in order to change it you have to first determine which stat the geom uses, then determine the arguments to that stat.

For example, here is the default histogram of Home_Value:

```{r}
#

#

```

can change it by passing the `binwidth` argument to the `stat_bin()` function:

```{r}
#

```


### Changing the transformation

Sometimes the default statistical transformation is not what you need. This is often the case with pre-summarized data:

```{r}
housing_sum <- 
  housing %>%
  group_by(State) %>%
  summarize(Home_Value_Mean = mean(Home_Value)) %>%
  ungroup()

head(housing_sum)
```

```{r, error=TRUE}
ggplot(housing_sum, aes(x=State, y=Home_Value_Mean)) + 
  geom_bar()
```

What is the problem with the previous plot? Basically we take binned and summarized data and ask ggplot to bin and summarize it again (remember, `geom_bar()` defaults to `stat = stat_count`; obviously this will not work. We can fix it by telling `geom_bar()` to use a different statistical transformation function:

```{r}
#

```


## Exercise 1

1.  Re-create a scatter plot with `CPI` on the x axis and `HDI` on the y axis (as you did in the previous exercise).
```{r}
## 
```

2.  Overlay a smoothing line on top of the scatter plot using `geom_smooth()`.
```{r}
## 
```

3.  Overlay a smoothing line on top of the scatter plot using `geom_smooth()`, but use a linear model for the predictions. Hint: see `?stat_smooth`.
```{r}
## 
```

4.  Overlay a smoothing line on top of the scatter plot using the default *loess* method for `geom_smooth()`, but make it less smooth. Hint: see `?stat_smooth` and the `span` argument.
```{r}
## 
```

5.  BONUS: Overlay a loess `(method = "loess")` smoothing line on top of the scatter plot using `geom_line()`. Hint: change the statistical transformation.
```{r}
## 
```


## Scales

### Controlling aesthetic mapping

Aesthetic mapping (i.e., with `aes()`) only says that a variable should be mapped to an aesthetic. It doesn't say *how* that should happen. For example, when mapping a variable to *shape* with `aes(shape = x)` you don't say *what* shapes should be used. Similarly, `aes(color = z)` doesn't say *what* colors should be used. Describing what colors/shapes/sizes etc. to use is done by modifying the corresponding *scale*. In `ggplot2` scales include

* position
* color and fill
* size
* shape
* line type

Scales are modified with a series of functions using a `scale_<aesthetic>_<type>` naming scheme. Try typing `scale_<tab>` to see a list of scale modification functions.

### Common scale arguments

The following arguments are common to most scales in `ggplot2`:

* **name:** the axis or legend title
* **limits:** the minimum and maximum of the scale
* **breaks:** the points along the scale where labels should appear
* **labels:** the labels that appear at each break

Specific scale functions may have additional arguments; for example, the `scale_color_continuous()` function has arguments `low` and `high` for setting the colors at the low and high end of the scale.

### Scale modification examples

Start by constructing a dotplot showing the distribution of home values by `Date` and `State`.

```{r}
p4 <- ggplot(housing, aes(x = State, y = Home_Price_Index)) + 
    geom_point(aes(color = Date), alpha = 0.5, size = 1.5,
               position = position_jitter(width = 0.25, height = 0))
```


Now modify the breaks for the color scales

```{r}
#

```

Next change the low and high values to blue and red:

```{r}
#

```
Now mute the colors:

```{r}
#

```


### Using different color scales

`ggplot2` has a wide variety of color scales; here is an example using `scale_color_gradient2()` to interpolate between three different colors.

```{r}
#

```


### Available scales

* Partial combination matrix of available scales

| `scale_`          | Types        |  Examples                   |
|:------------------|:-------------|:----------------------------|
| `scale_color_`    | `identity`   | `scale_fill_continuous()`   |
| `scale_fill_`     | `manual`     | `scale_color_discrete()`    |
| `scale_size_`     | `continuous` | `scale_size_manual()`       |
|                   | `discrete`   | `scale_size_discrete()`     |
|                   |              |                             |
| `scale_shape_`    | `discrete`   | `scale_shape_discrete()`    |
| `scale_linetype_` | `identity`   | `scale_shape_manual()`      |
|                   | `manual`     | `scale_linetype_discrete()` |
|                   |              |                             |
| `scale_x_`        | `continuous` | `scale_x_continuous()`      |
| `scale_y_`        | `discrete`   | `scale_y_discrete()`        |
|                   | `reverse`    | `scale_x_log()`             |
|                   | `log`        | `scale_y_reverse()`         |
|                   | `date`       | `scale_x_date()`            |
|                   | `datetime`   | `scale_y_datetime()`        |

Note that in RStudio you can type `scale_` followed by `tab` to get the whole list of available scales. 
For a complete list of available scales see <https://ggplot2.tidyverse.org/reference/>


## Exercise 2

1.  Create a scatter plot with `CPI` on the x axis and `HDI` on the y axis. Color the points to indicate `Region`.
```{r}
## 
```

2.  Modify the x, y, and color scales so that they have more easily-understood names (e.g., spell out "Human development Index" instead of `HDI`). Hint: see `?scale_x_discrete`.
```{r}
## 
```

3.  Modify the color scale to use specific values of your choosing. Hint: see `?scale_color_manual` and <https://www.color-hex.com/>.
```{r}
## 
```


## Faceting

### What is faceting?

* Faceting is `ggplot2` parlance for **small multiples**
* The idea is to create separate graphs for subsets of data
* `ggplot2` offers two functions for creating small multiples:
    1.  `facet_wrap()`: define subsets as the levels of a single grouping variable
    2.  `facet_grid()`: define subsets as the crossing of two grouping variables
* Facilitates comparison among plots, not just of geoms within a plot

### What is the trend in housing prices in each state?

* Start by using a technique we already know; map `State` to color:

```{r}
#

#

```

There are two problems here; there are too many states to distinguish each one by color, and the lines obscure one another.

### Faceting to the rescue

We can remedy the deficiencies of the previous plot by faceting by `State` rather than mapping `State` to color.

```{r}
#

#

```


## Themes

### What are themes?

The `ggplot2` theme system handles non-data plot elements such as:

* Axis label properties (e.g., font, size, color, etc.)
* Plot background
* Facet label background
* Legend appearance

Built-in themes include:

* `theme_gray()` (default)
* `theme_bw()`
* `theme_classic()`

```{r}
#

```

```{r}
#

```
You can see a list of available built-in themes here <https://ggplot2.tidyverse.org/reference/>


### Overriding theme defaults

Specific theme elements can be overridden using `theme()`. For example:

```{r}
#

```

All theme options are documented in `?theme`. We can also see the
existing default values using:

```{r, eval=FALSE}
theme_get()
```

### Creating & saving new themes

You can create new themes, as in the following example:

```{r}
theme_new <- theme_bw() +
  theme(plot.background = element_rect(size = 1, color = "blue", fill = "black"),
        text = element_text(size = 12, family = "Serif", color = "ivory"),
        axis.text.y = element_text(colour = "purple"),
        axis.text.x = element_text(colour = "red"),
        panel.background = element_rect(fill = "pink"),
        strip.background = element_rect(fill = muted("orange")))

p5 + theme_new
```

## Saving plots

We can save a plot to either a vector (e.g., pdf, eps, ps, svg) 
or raster (e.g., jpg, png, tiff, bmp, wmf) graphics file using
the `ggsave()` function:

```{r, eval=FALSE}
#

```


## The #1 FAQ

### Map aesthetic to different columns

The most frequently asked question goes something like this: *I have two variables in my data.frame, and I'd like to plot them as separate points, with different color depending on which variable it is. How do I do that?*

**Wrong**

```{r}
housing_byyear <- 
  housing %>%
  group_by(Date) %>%
  summarize(Home_Value_Mean = mean(Home_Value),
            Land_Value_Mean = mean(Land_Value)) %>%
  ungroup()

ggplot(housing_byyear, aes(x=Date)) +
  geom_line(aes(y=Home_Value_Mean), color="red") +
  geom_line(aes(y=Land_Value_Mean), color="blue")
```

**Right**

```{r}
home_land_byyear <- gather(housing_byyear,
                           value = "value",
                           key = "type",
                           Home_Value_Mean, Land_Value_Mean)

ggplot(home_land_byyear, aes(x=Date, y=value, color=type)) +
  geom_line()
```


## Exercise 3

For this exercise, we're going to use the built-in `midwest` dataset:

```{r}
data("midwest", package = "ggplot2")
head(midwest)
```

1.  Create a scatter plot with `area` on the x axis and the log of `poptotal` on the y axis. 
```{r}
## 
```

2.  Within the `geom_point()` call, map color to `state`, map size to the log of `popdensity`, and fix transparency (`alpha`) to 0.3.
```{r}
## 
```

3.  Add a smoother and turn off plotting the confidence interval. Hint: see the `se` argument to `geom_smooth()`.
```{r}
## 
```

4.  Facet the plot by `state`. Set the `scales` argument to `facet_wrap()` to allow separate ranges for the x-axis.
```{r}
## 
```

5.  Change the default color scale to use the discrete `RColorBrewer` palette called `Set1`. Hint: see `?scale_color_brewer`.
```{r}
## 
```

6.  BONUS: Change the default theme to `theme_bw()` and modify it so that the axis text and facet label background are blue. Hint: see `?theme` and especially `axis.text` and `strip.background`.
```{r}
## 
```




## Wrap-up

### Feedback

These workshops are a work in progress, please provide any feedback to: help@iq.harvard.edu

### Resources

* IQSS 
    + Workshops: <https://dss.iq.harvard.edu/workshop-materials>
    + Data Science Services: <https://dss.iq.harvard.edu/>
    + Research Computing Environment: <https://iqss.github.io/dss-rce/>

* HBS
    + Research Computing Services workshops: <https://training.rcs.hbs.org/workshops>
    + Other HBS RCS resources: <https://training.rcs.hbs.org/workshop-materials>
    + RCS consulting email: <mailto:research@hbs.edu>
    
* ggplot2
    + Reference: <https://ggplot2.tidyverse.org/reference/>
    + Cheatsheets: <https://rstudio.com/wp-content/uploads/2019/01/Cheatsheets_2019.pdf>
    + Examples: <http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html>
    + Tutorial: <https://uc-r.github.io/ggplot_intro>
    + Mailing list: <http://groups.google.com/group/ggplot2>
    + Wiki: <https://github.com/hadley/ggplot2/wiki>
    + Website: <http://had.co.nz/ggplot2/>
    + StackOverflow: <http://stackoverflow.com/questions/tagged/ggplot>

