

## Exercise solutions

### Ex 0: prototype

1.  Create a scatter plot with `CPI` on the x axis and `HDI` on the y axis.

```{r}
ggplot(dat, aes(x = CPI, y = HDI)) +
  geom_point()
```

2.  Color the points in the previous plot blue.

```{r}
ggplot(dat, aes(x = CPI, y = HDI)) +
  geom_point(color = "blue")
```

3.  Map the color of the the points to `Region`.

```{r}
ggplot(dat, aes(x = CPI, y = HDI)) +
  geom_point(aes(color = Region))
```

4.  Keeping color mapped to `Region`, make the points bigger by setting size to 2.

```{r}
ggplot(dat, aes(x = CPI, y = HDI)) +
  geom_point(aes(color = Region), size = 2)
```

5.  Keeping color mapped to `Region`, map the size of the points to `HDI_Rank`.

```{r}
ggplot(dat, aes(x = CPI, y = HDI)) +
geom_point(aes(color = Region, size =  HDI_Rank))
```


### Ex 1: prototype

1.  Re-create a scatter plot with `CPI` on the x axis and `HDI` on the y axis (as you did in the previous exercise).

```{r}
ggplot(dat, aes(x = CPI, y = HDI)) +
  geom_point()
```

2.  Overlay a smoothing line on top of the scatter plot using `geom_smooth()`.

```{r}
ggplot(dat, aes(x = CPI, y = HDI)) +
  geom_point() +
  geom_smooth()
```

3.  Make the smoothing line in `geom_smooth()` less smooth. Hint: see `?loess`.

```{r}
ggplot(dat, aes(x = CPI, y = HDI)) +
  geom_point() +
  geom_smooth(span = .4)
```

4.  Change the smoothing line in `geom_smooth()` to use a linear model for the predictions. Hint: see `?stat_smooth`.

```{r}
ggplot(dat, aes(x = CPI, y = HDI)) +
  geom_point() +
  geom_smooth(method = "lm")
```

5.  BONUS: Overlay a loess `(method = "loess")` smoothing line on top of the scatter plot using `geom_line()`. Hint: change the statistical transformation.

```{r}
ggplot(dat, aes(x = CPI, y = HDI)) +
  geom_point() +
  geom_line(stat = "smooth", method = "loess")
```



### Ex 2: prototype

1.  Create a scatter plot with `CPI` on the x axis and `HDI` on the y axis. Color the points to indicate `Region`.

```{r}
ggplot(dat, aes(x = CPI, y = HDI, color = Region)) +
  geom_point()
```

2.  Modify the x, y, and color scales so that they have more easily-understood names (e.g., spell out "Human development Index" instead of `HDI`).

```{r}
ggplot(dat, aes(x = CPI, y = HDI, color = Region)) +
geom_point() +
scale_x_continuous(name = "Corruption Perception Index") +
scale_y_continuous(name = "Human Development Index") +
scale_color_discrete(name = "Region of the world")
```

3.  Modify the color scale to use specific values of your choosing. Hint: see `?scale_color_manual`. NOTE: you can specify color by name (e.g., "blue") or by "Hex value" --- see <https://www.color-hex.com/>.

```{r}
ggplot(dat, aes(x = CPI, y = HDI, color = Region)) +
geom_point() +
scale_x_continuous(name = "Corruption Perception Index") +
scale_y_continuous(name = "Human Development Index") +
  scale_color_manual(name = "Region of the world",
                     values = c("red", "green", "blue", "orange", "grey", "brown"))
```


### Ex 3: prototype

1.  Create a scatter plot with `area` on the x axis and the log of `poptotal` on the y axis. 

```{r}
p6 <- ggplot(midwest, aes(x=area, y=log(poptotal))) 
p6 + geom_point() 
```

2.  Within the `geom_point()` call, map color to `state`, map size to the log of `popdensity`, and fix transparency (`alpha`) to 0.3.

```{r}
p6 <- p6 + geom_point(aes(color=state, size=log(popdensity)), alpha = 0.3) 
```

3.  Add a smoother and turn off plotting the confidence interval. Hint: see the `se` argument to `geom_smooth()`.

```{r}
p6 <- p6 + geom_smooth(method="loess", se=FALSE) 
```

4.  Facet the plot by `state`. Set the `scales` argument to `facet_wrap()` to allow separate ranges for the x-axis.

```{r}
p6 <- p6 + facet_wrap(~ state, scales = "free_x")
```

5.  Change the default color scale to use the discrete `RColorBrewer` palette called `Set1`. Hint: see `?scale_color_brewer`.

```{r}
p6 <- p6 + scale_color_brewer(palette = "Set1")
```

6.  BONUS: Change the default theme to `theme_bw()` and modify it so that the axis text and facet label background are blue. Hint: see `?theme` and especially `axis.text` and `strip.background`.

```{r}
p6 <- p6 + theme_bw() +
    theme(axis.title = element_text(color = "blue", face = "bold"),
          strip.background = element_rect(fill = "yellow"))
```

Here's the complete code for the Exercise 3 plot:

```{r}
p6 <- ggplot(midwest, aes(x=area, y=log(poptotal))) +
    geom_point(aes(color=state, size=log(popdensity)), alpha = 0.3) +
    geom_smooth(method="loess", se=FALSE) +
    facet_wrap(~ state, scales = "free_x") +
    scale_color_brewer(palette = "Set1") +
    theme_bw() +
    theme(axis.title = element_text(color = "blue", face = "bold"),
         strip.background = element_rect(fill = "yellow"))
```

