
## Exercise solutions

### Ex 0: prototype

1. 2 plus 2

    ```{r}
    2 + 2
    # or
    sum(2, 2)
    ```

2. square root of 10

    ```{r}
    sqrt(10)
    # or
    10^(1/2)
    ```

3. Find "An Introduction to R".

    ```{r}
    # Go to the main help page by running 'help.start() or using the GUI
    # menu, find and click on the link to "An Introduction to R".
    ```


### Ex 1: prototype

1. Open the `read_csv()` help page to determine how to use it to read in data.

    ```{r, eval=FALSE}
    ?read_csv
    ```

2. Read the baby names data using the `read_csv()` function and assign the result with the name `baby_names`.

    ```{r}
    baby_names <- read_csv("babyNames.csv")
    ```

3. BONUS (optional): Save the `baby_names` data as a Stata data set `babynames.dta` and as an R data set `babynames.rds`.
 
    ```{r, eval=FALSE} 
    write_dta(baby_names, version = 15, path = “babynames.dta”)
     
    write_rds(baby_names, file = “babynames.rds”)
    ```


### Ex 2.1: prototype

1. Use `filter` to extract data for your name (or another name of your choice).

    ```{r}
    baby_names_george <- filter(baby_names, Name == "George")
    ```

2. Arrange the data you produced in step 1 above by `Count`. In which year was the name most popular?

    ```{r}
    arrange(baby_names_george, desc(Count))
    ```

3. BONUS (optional): Filter the data to extract _only_ the row containing the most popular boys name in 1999.

    ```{r}
    baby_names_boys_1999 <- filter(baby_names,
                                   Year == 1999 & Sex == "Boys")
    ```

    ```{r}
    filter(baby_names_boys_1999, Count == max(Count))
    ```

### Ex 2.2: prototype

Rewrite the solution to Exercise 2.1 using pipes.

```{r}
baby_names %>% 
    filter(Name == "George") %>%
    arrange(desc(Count))
```


### Ex 3: prototype

1. Use `filter()` to extract data for your name (same as previous exercise).  

    ```{r}
    baby_names_george <- filter(baby_names, Name == "George")
    ```

2. Plot the data you produced in step 1 above, with `Year` on the x-axis and `Count` on the y-axis.

    ```{r}
    qplot(x = Year, y = Count, data = baby_names_george)
    ```

3. Adjust the plot so that is shows boys and girls in different colors.

    ```{r}
    qplot(x = Year, y = Count, color = Sex, data = baby_names_george)
    ```

4. BONUS (Optional): Adjust the plot to use lines instead of points.

    ```{r}
    qplot(x = Year, y = Count, color = Sex, data = baby_names_george, geom = "line")
    ```


### Ex 4: prototype

1. Use `mutate()` and `group_by()` to create a column named `Proportion` where `Proportion = Count/sum(Count)` for each `Year X Sex` group.

    ```{r}
    baby_names <- 
      baby_names %>%
      group_by(Year, Sex) %>%
      mutate(Proportion = Count/sum(Count)) %>%
      ungroup()

    head(baby_names) 
    ```

2. Use `mutate()` and `group_by()` to create a column named `Rank` where `Rank = rank(desc(Count))` for each `Year X Sex` group.

    ```{r}
    baby_names <- 
      baby_names %>%
      group_by(Year, Sex) %>%
      mutate(Rank = rank(desc(Count))) %>%
      ungroup()

    head(baby_names)
    ```

3. Filter the baby names data to display only the most popular name for each `Year X Sex` group.

    ```{r}
    top1 <- 
      baby_names %>%
      filter(Rank == 1) %>%
      select(Year, Name, Sex, Proportion)

    head(top1)
    ```

4. Plot the data produced in step 3, putting `Year` on the x-axis and `Proportion` on the y-axis. How has the proportion of babies given the most popular name changed over time?

    ```{r}
    qplot(x = Year, 
          y = Proportion, 
          color = Sex, 
          data = top1, 
          geom = "line")
    ```

5. BONUS (optional): Which names are the most popular for both boys and girls?

    ```{r}
    girls_and_boys <- inner_join(filter(baby_names, Sex == "Boys"), 
                                filter(baby_names, Sex == "Girls"),
                                by = c("Year", "Name"))

    girls_and_boys <- mutate(girls_and_boys,
                            Product = Count.x * Count.y,
                            Rank = rank(desc(Product)))

    filter(girls_and_boys, Rank == 1)
    ```


### Ex 5: prototype

1. Filter the baby_names data, retaining only the 10 most popular girl and boy names for each year.

    ```{r}
    most_popular <- 
      baby_names %>% 
      group_by(Year, Sex) %>%
      filter(Rank <= 10)

    head(most_popular, n = 10)
    ```

2. Summarize the data produced in step one to calculate the total Proportion of boys and girls given one of the top 10 names each year.

    ```{r}
    top10 <- 
      most_popular %>% # it is already grouped by Year and Sex
      summarize(TotalProportion = sum(Proportion))
    ```

3. Plot the data produced in step 2, with year on the x-axis and total proportion on the y axis. Color by `Sex`.

    ```{r}
    qplot(x = Year, 
          y = TotalProportion, 
          color = Sex,
          data = top10,
          geom = "line")
    ```

