
## Exercise solutions

### Ex 0: prototype

>     Locate the files named `1996boys_tcm77-254026.xlsx` and 
>     `2015boysnamesfinal.xlsx` and open them separately in a 
>      spreadsheet program. 
>
>     (If you don't have a spreadsheet program installed on
>     your computer you can download one from
>     https://www.libreoffice.org/download/download/). 
>
>     What issues can you identify that might make working
>     with these data difficult?
>
>     In what ways is the format different between the two files?

1.  Multiple Excel sheets in each file, each with a different name, but each file contains a `Table 1`.
2.  The data does not start on row one. Headers are on row 7, followed by a blank line, followed by the actual data.
3.  The data is stored in an inconvenient way, with ranks 1-50 in the first set of columns and ranks 51-100 in a second set of columns.
4. The second worksheet `2015boysnamesfinal.xlsx` contains extra columns between the data of interest, resulting in the second set of columns (ranks 51-100) being placed in a different position.
5.  The year from which the data comes is only reported in the Excel file name, not within the data itself.
6.  There are notes below the data.

These differences will make it more difficult to automate
re-arranging the data since we have to write code that can handle
different input formats.


### Ex 1: prototype

1.  Write a function that takes a file name as an argument and reads the worksheet containing "Table 1" from that file.

```{r} 
read_boys_names <- function(file, sheet_name) {
  read_excel(
    path = file,
    sheet = get_data_sheet_name(file, term = sheet_name),
    skip = 6
  )
}
```

2.  Test your function by using it to read *one* of the boys names Excel files.

```{r}
read_boys_names(boy_file_names[1], sheet_name = "Table 1") %>% glimpse()
```

3.  Use the `map()` function to read data from all the Excel files, using the function you wrote in step 1.

```{r}
boysNames <- map(boy_file_names, read_boys_names, sheet_name = "Table 1")
```

### Ex 2: prototype

1.  Write a function that takes a data frame as an argument and returns a modified version, which keeps only columns that include the strings `Name` and `Count` in the column names. HINT: see the `?matches` function.

```{r}
  namecount <- function(data) {
      select(data, matches("Name|Count"))
  }
```

2.  Test your function on the first data frame in the list of boys names data.

```{r}
  namecount(boysNames[[1]])
```

3.  Use the `map()` function to each data frame in the list of boys names data.

```{r}
  boysNames <- map(boysNames, namecount)
```

### Ex 3: prototype

1.  Create a new function called `cleanupNamesData` that:

```{r}
cleanupNamesData <- function(file){

  # subset data to include only those columns that include the term `Name` and `Count`
  subsetted_file <- file %>%
    select(matches("Name|Count")) %>%
    drop_na()

  # subset two separate data frames, with first and second set of `Name` and `Count` columns 
  first_columns <- select(subsetted_file, Name = Name...2, Count = Count...3) 

  second_columns <- select(subsetted_file, Name = matches("Name...6|Name...7|Name...8"),
                                           Count = matches("Count...7|Count...8|Count...9"))

  # append the two datasets
  bind_rows(first_columns, second_columns)
}


## test it out on the second data frame in the list
boysNames[[2]] %>% glimpse() # before cleanup
boysNames[[2]] %>% cleanupNamesData() %>% glimpse() # after cleanup
```

2.  Your task now is to use the `map()` function to apply each of these transformations to all the elements in `boysNames`. 

```{r}
boysNames <- map(boysNames, cleanupNamesData)
```

### Ex 4: prototype

1.  Turn the list of boys names data frames into a single data frame.

```{r}
boysNames <- bind_rows(boysNames)
glimpse(boysNames)
```

2.  Create a new directory called `all` within `dataSets` and write the data to a `.csv` file. HINT: see the `?dir.create` and `?write_csv` functions.

```{r, eval=FALSE}
dir.create("dataSets/all")

write_csv(boysNames, "dataSets/all/boys_names.csv")
```

3.  What were the five most popular names in 2013?   

```{r}
boysNames %>% 
  filter(Year == 2013) %>%
  arrange(desc(Count)) %>%
  head()
```

4.  How has the popularity of the name "ANDREW" changed over time?

```{r}
andrew <- filter(boysNames, Name == "ANDREW")

ggplot(andrew, aes(x = Year, y = Count)) +
    geom_line() +
    ggtitle("Popularity of Andrew, over time")
```


## Complete code

1.  Code for Section 1: Reading data from multiple Excel worksheets into R data frames
```{r}
boy_file_names <- list.files("dataSets/boys", full.names = TRUE)

get_data_sheet_name <- function(file, term){
  excel_sheets(file) %>% str_subset(pattern = term)
}

read_boys_names <- function(file, sheet_name) {
  read_excel(
    path = file,
    sheet = get_data_sheet_name(file, term = sheet_name),
    skip = 6
  )
}

boysNames <- map(boy_file_names, read_boys_names, sheet_name = "Table 1")
```

2.  Code for Section 2: Clean up data within each R data frame
```{r}
cleanupNamesData <- function(file){
  # subset data to include only those columns that include the term `Name` and `Count`
  subsetted_file <- file %>%
    select(matches("Name|Count")) %>%
    drop_na()
  # subset two separate data frames, with first and second set of `Name` and `Count` columns 
  first_columns <- select(subsetted_file, Name = Name...2, Count = Count...3) 
  second_columns <- select(subsetted_file, Name = matches("Name...6|Name...7|Name...8"),
                                           Count = matches("Count...7|Count...8|Count...9"))
  # append the two datasets
  bind_rows(first_columns, second_columns)
}

boysNames <- map(boysNames, cleanupNamesData)
```

3.  Code for Section 3: Organize the data into one large data frame and store it
```{r}
Years <- str_extract(boy_file_names, pattern = "[0-9]{4}")

names(boysNames) <- Years

boysNames <- imap(boysNames, ~ mutate(.x, Year = as.integer(.y)))

boysNames <- bind_rows(boysNames)
```


